Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-22 paper/ms63.tex
adding some new citations and orcid
another orcid
adding modified text from referee report

\author{Jason Curtis}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

@@

\author{Jacqueline K. Faherty}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

@@

The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.
There are many other empirical gyrochronology models we could have chosen to
compare with our data.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data here.
The \citet{barnes2010} model has a similar period-\teff\ relation to the
Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
Praesepe roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
in its current form does not have the flexibility to capture the mass and
time-variable rotational evolution revealed by the data explored here.

The lack of flexibility in these empirical models should not be seen as a
failing of gyrochronology.
With a paucity of old calibration stars, the only possible approach has been
to fit simple relations to young stars and extrapolate to old ages.
Naturally however, in this era of precise space-based photometry, new data are
revealing flaws in our old models (which no one ever thought were perfect),
and are providing opportunities to improve them.
}

>>>>

\author[0000-0002-2792-134X]{Jason L. Curtis}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

@@

\author[0000-0001-6251-0573]{Jacqueline K. Faherty}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

@@

The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.
Most semi-physical spin-down models predict that, once the rotation periods of
stars converge to a tight sequence, they approximately spin down with a common
braking index \citep[e.g., Fig. 5 in][]{gallet2015}.
While some semi-physical models do not predict that stars have fully converged
by the age of Praesepe, the observations suggest that they have
\citep{douglas2019}, in which case, projecting the Praesepe sequence forward
with a single braking index that is common to all stars and constant in time
is appropriate.
However, observations of low-mass stars in older clusters \citep[\eg, NGC
6811, NGC 752, and Ruprecht 147][Curtis \etal, submitted]{curtis2019,
agueros2018} demonstrate that the assumptions adopted in those models must be
invalid, or inaccurate in some quantitative way (for example, the
core-envelope coupling timescale is inaccurately calibrated).

There are many other empirical gyrochronology models we could have chosen to
compare with our data.
For example, the \citet{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015} relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ ($\approx$
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data here.
The \citet{barnes2010} model has a similar period-\teff\ relation to the
Praesepe-based \citet{angus2019} model, because the period-\teff\ relation of
Praesepe roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010} model
in its current form does not have the flexibility to capture the mass and
time-variable rotational evolution revealed by the data explored here.

% The lack of flexibility in these empirical models should not be seen as a
% failing of gyrochronology.
% With a paucity of old calibration stars, the only possible approach has been
% to fit simple relations to young stars and extrapolate to old ages.
% Naturally however, in this era of precise space-based photometry, new data are
% revealing flaws in our old models (which no one ever thought were perfect),
% and are providing opportunities to improve them.

It is not surprising that the \citet{angus2019} model, or other gyrochronology
models do not reproduce the data, since tensions with
\citet{skumanich1972}-type spin-down have been revealed whenever new rotation
periods of benchmark stars have become available.
With the influx of new stellar rotation periods provided by \kepler/\ktwo\ and
\tess, our understanding of spin-down is changing rapidly.
New data are revealing flaws in not just the calibration of but also the
functional forms of old models.
We have shown that velocity dispersion data provides support for more
sophisticated models incorporating additional physical effects \citep[\eg,
core-envelope coupling][]{spada2019} over simple empirical models.
Alternatively, more flexible empirical frameworks, like a Gaussian process
model, could capture everything we now know about stellar spin-down.
}
